Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-29 vertical-thermal-review.Rmd
cutting words from Tom's suggestion
Please let me know what you think of these changes Krista!

Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. 
The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021]. 
The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019]. 
This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

Here, we review how the biophysical environment and plant form and function vary across the vertical gradient in forests. 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and thermal tolerance;
(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, and respiration) and its thermal sensitivity; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.

@@

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005]. 
In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021]. 
Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers. 
Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).

@@

Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019]. 
It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007-->; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019-->; @zellweger_seasonal_2019; <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->]. 
Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021]. 
Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig. `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987-->; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.

>>>>

Forest responses and feedbacks to climate change will critically influence Earth's furture climate.
Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. 
Warming may increase or decreased tree growth and forest carbon sequestration on balance, with decreases being more common globally [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021]. 
The resulting feedbacks to climate and carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

Much uncertainty arises because effects of warming on trees depend on size and crown position. 
Forests are vertically and horizontally stratified, and canopies moderate climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019]. 
The resulting vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
Although these vertical gradients strongly influence plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

Here, we review vertical gradients in the biophysical environment and plant form and function in forests, focusing on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and thermal tolerance;
(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, and respiration) and its thermal sensitivity; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change and how they scale across space and time.

@@

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005]. 
In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].

In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers. 
Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).

@@

Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019]. 
It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->]. 
Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021]. 
Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig. `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.
